\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-07-29 writing/tex/draft_introduction.tex
Finished the abstract earlier today. With that, the first draft is done!
Made all the changes that I wrote down on the printed out draft. Up next: throwing it through Docs to check grammar, and having it read by one or two friends!

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (VT), gained state of the art performance on common learning benchmarks, including the ImageNet dataset\footnote{At the time of writing the VT \texttt{CoCa} is state of the art, according to \url{https://paperswithcode.com/sota/image-classification-on-imagenet}.}.

@@

\subsection{Vision Transformers} \label{vits}
\paragraph{Architecture}
VTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{encoder stack} of the \textit{Transformer} architecture \citep{vaswani2017attention}. This stack takes a sequence of vectors as input, and is composed of alternating \textit{multi-headed self-attention} layers and position-wise MLP-type layers -- position-wise meaning the MLP can take only a single position (constituent vector) as input at a time, and is shared between all positions.

@@

The original Transformer architecture was designed to solve problems in \textit{natural language processing} (a field it now dominates). To make it work for CV applications, VTs divide an image into square patches, which get treated as tokens (words). Learned embeddings convert these into a sequence of vectors, to which 1D positional encodings are added\footnote{Do mind that VTs have no built-in knowledge of an image's 2D topology.}. A learned $[class]$ vector gets prepended to the sequence, and its representation at the end of the encoder stack is the final output, which can then be used for classification, segmentation, etc.

@@

Moreover, weight sharing applies as much to VTs as it does to CNNs. In VTs, each position goes through the same computational pipelines of being multiplied by 3 projection matrices, followed by a weighted sum calculation, followed by being fed through an MLP-type layer\footnote{Neglecting skip connections and normalization, as this is a comparison on an intuitive level.}. In CNNs these pipelines take the form of convolution kernels.

\paragraph{A limitation of VTs}
While VTs are promising; the fact that they are not tailored to CV tasks, makes that they lack a so called image specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a VT first has to learn what an image is, before it can move on to the learning task at hand. The result is that they require substantially more training data (and time) than CNNs do.

@@

In practice, many TL techniques exist. The example given in section \ref{vits} can be described as \textit{fine-tuning} (FT) TL. Here, a pre-trained model is used as a starting point, and then trained on the target task -- possibly with different hyperparameters, and often with a new final layer that fits the task. \citeauthor{yosinski2014transferable} (\citeyear{yosinski2014transferable}) studied transferability properties of CNNs pre-trained on ImageNet, and suggested that FT results in improved performance and generalizability compared to training from scratch. Moreover, while transferability decreased as source and target task became more dissimilar, utilizing TL still seemed to be preferred.

@@

A different TL technique is called \textit{off-the-shelf} (OTS) TL. It is similar to FT, except that all but the final layer is kept frozen during training. \citeauthor{sharif2014cnn} (\citeyear{sharif2014cnn}) found that features extracted by a CNN pre-trained on ImageNet were general enough for OTS TL to work. More specifically, these OTS CNNs showed better performance than the (fully trained non-CNN) state of the art at the time. Besides reducing computational cost of training, a benefit of OTS TL is that it can prevent overfitting in cases where FT would not \citep{yosinski2014transferable}.

@@

\subsection{Related work and the current study} \label{related_work}
As mentioned, the paper that first introduced VTs already demonstrated they benefit from TL \citep{dosovitskiy2020image}, and more has been discovered about VTs' TL properties since then. \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) investigated whether TL allowed VTs to replace CNNs in Medical Imaging. This domain is characterized by small datasets, so to overcome the inductive bias problem described earlier, TL is imperative. It was found that VTs benefit more from FT TL than CNNs do, and that VTs pre-trained on ImageNet perform on par CNNs.

@@

Another `shortcomming' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not necessarily related to TL as defined in this paper, however.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current study will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempts to find a methodology that is recommended in these circumstances. It will, however, restrict itself to art classification tasks. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

%TODO: talk about datasets small enough to be hand labelled by one person
%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

More formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilizing TL allow VTs to perform on par with CNNs on common art classification problems, when compute resources are limited to a single GPU?}\end{quote}

>>>>

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (VT), gained state of the art performance on common learning benchmarks, including the ImageNet dataset \citep{deng2009imagenet}\footnote{At the time of writing, the VT \texttt{CoCa} is state of the art, according to \url{https://paperswithcode.com/sota/image-classification-on-imagenet}.}.

@@

\subsection{Vision Transformers} \label{vits}
\paragraph{Architecture}
VTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{encoder stack} of the \textit{Transformer} architecture \citep{vaswani2017attention}. This stack takes a sequence of vectors as input, and is composed of alternating \textit{multi-headed self-attention} layers and position-wise MLP-type layers -- position-wise meaning the MLP's input is the size of a single position (constituent vector), and is shared between all positions.

@@

The original Transformer architecture was designed to solve problems in \textit{natural language processing} (a field it now dominates). To make it work for CV applications, VTs divide an image into square patches, which get treated as tokens (words). Learned embeddings convert these into a sequence of vectors, to which 1D positional embeddings are added\footnote{Do mind that VTs have no built-in knowledge of an image's topology, as the architecture ignores the 2D layout.}. A learned $[class]$ vector gets prepended to the sequence, and its representation at the end of the encoder stack is the final output, which can then be used for classification, segmentation, etc.

@@

Moreover, weight sharing applies as much to VTs as it does to CNNs. In VTs, each position goes through the same computational pipeline of being multiplied by 3 projection matrices, followed by a weighted sum calculation, followed by being fed through an MLP-type layer\footnote{Neglecting skip connections, normalization and multi-headedness, as this is a comparison on an intuitive level.}. In CNNs these pipelines take the form of convolution kernels.

\paragraph{A limitation of VTs}
While VTs are promising; the fact that they are not specifically designed for CV tasks, makes that they lack a so called image specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a VT first has to learn what an image is, before it can move on to the learning task at hand\footnote{Recall, for example, that the architecture ignores the 2D layout of an image, and hence has to learn about this through training.}. The result is that they require substantially more training data (and time) than CNNs do.

@@

In practice, many TL techniques exist. The example given in section \ref{vits} can be described as \textit{fine-tuning} (FT). Here, a pre-trained model is used as a starting point, and then trained on the target task -- possibly with different hyperparameters, and often with a new final layer that fits the task. \citeauthor{yosinski2014transferable} (\citeyear{yosinski2014transferable}) studied transferability properties of CNNs pre-trained on ImageNet, and suggested that FT results in improved performance and generalizability compared to training from scratch. Moreover, while transferability decreased as source and target task became more dissimilar, utilizing TL still seemed to be preferred.

@@

Another TL technique is called \textit{off-the-shelf} (OTS) learning. It is similar to FT, except that all but the final layer is kept frozen during training. \citeauthor{sharif2014cnn} (\citeyear{sharif2014cnn}) found that features extracted by a CNN pre-trained on ImageNet were general enough for OTS learning to work. More specifically, these OTS CNNs showed better performance than the (fully trained non-CNN) state of the art at the time. Besides reducing computational cost of training, a benefit of OTS learning is that it can prevent overfitting in cases where FT would not \citep{yosinski2014transferable}.

@@

\subsection{Related work and the current study} \label{related_work}
As mentioned, the paper that first introduced VTs already demonstrated they benefit from TL \citep{dosovitskiy2020image}, and more has been discovered about VTs' TL properties since then. \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) investigated whether TL allowed VTs to replace CNNs in Medical Imaging. This domain is characterized by small datasets, so to overcome the inductive bias problem described earlier, TL is imperative. It was found that VTs benefit more from FT than CNNs do, and that VTs pre-trained on ImageNet perform on par with CNNs after this type of TL.

@@

Another `shortcomming' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterised by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not related to FT or OTS learning, of course.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current thesis will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when resources are limited, and attempts to find a methodology that is recommended in these circumstances. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

Limited resources, in this case, primarily refers to the amount of available training data. Datasets should be so small that one can reasonable assume they were hand labelled by a single person. In addition, it also refers to compute resources, which should be restricted to consumer-grade hardware. In practice, of course, the small datasets already ensure that this restriction is met, because these naturally allow such hardware to suffice for learning.

Furthermore, this thesis restricts itself to art classification tasks specifically. These tasks fit in well with the described objectives, since the amount of effort needed to create and digitise an art collection often makes for small datasets. Moreover, they provide an interesting TL challenge, as knowledge might have to be carried over from natural tot non-natural images.
%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

To put all of this more formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilising TL allow VTs to perform on par with CNNs on common art classification problems, when datasets contain 10,000 images or less?}\end{quote}

### 2022-07-29 writing/tex/draft_methods.tex
Made all the changes that I wrote down on the printed out draft. Up next: throwing it through Docs to check grammar, and having it read by one or two friends!

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pre-trained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that FT resulted in substantially better performance than either training from scratch or using an OTS learning scheme.

@@

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K, while the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used for the target tasks. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks extracted from it are: (1) \textit{Type classification}, containing labels as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with labels as `paper', `porcelain', `silver', etc.; and finally, (3) \textit{Artist classification}, where the model has to predict who the creator is. Figure \ref{methods:eg} shows labelled examples for all three tasks.

The full dataset contains 112,039 images, and allows for multiple or zero labels per classification task. Samples are excluded from a task if they contain no labels for it. Samples containing more than one label are excluded as well, since analysis showed that in all cases this is true for only a small portion of the dataset.% Note that this last point is different from \citeauthor{sabatelli2018deep}, which took the first occurring class in these cases.

@@

All datasets are split up into 80\%, 10\%, 10\% training, validating and testing sets. For robustness, five subsets are randomly generated per classification task, such that each experiment can be run five times. The average sample overlap between two of the five datasets is shown in table \ref{methods:datasets} as well.

Besides the tasks described above, table \ref{methods:datasets} shows an experiment called \textit{Scaling}. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are gradually shrunk. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\% and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure the distribution remains the same. Finally, also here 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

To start off with CNNs, \texttt{ResNet50} \citep{he2016deep} and \texttt{VGG19} \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for respectively FT and OTS TL. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used \texttt{ResNet50} too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used \texttt{ResNet101} and \texttt{ResNet152}. To also add some more recent CNNs, \texttt{ConvNext} \citep{liu2022convnet} -- a purely CNN-based model inspired by VTs' recent success -- and \texttt{EfficientNetV2} \citep{tan2021efficientnetv2} are included.

For VTs, versions with a 16x16 patch size are chosen. The first model is the original VT \citep{dosovitskiy2020image}, which will be referred to as \texttt{ViT}. Next is \texttt{Swin} \citep{liu2021swin}, because it showed promising performance in \citeauthor{zhou2021convnets}. The final two are \texttt{DeiT} and \texttt{BeiT}. The small version of \texttt{DeiT} was used in \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}), because it is similar in size to \texttt{ResNet50}, to which it was compared. In this work, however, also a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.

@@

All VT-based models have a bit over 86 million parameters. \texttt{ConvNext} is of similar size, with 88.6 million parameters. \texttt{ResNet50} and \texttt{EfficientNetV2} are smaller, with 25.6 and 13.6 million, respectively, while \texttt{VGG19} is the largest model, with 143.7 million parameters. It should be mentioned that for both OTS TL and FT, preliminary experiments were performed with `small' and `tiny' models. The results of these suggested that similar conclusions could have been drawn if these were used instead.

@@

For OTS TL, the Adam optimizer is used with standard \texttt{PyTorch} parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for VT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32 GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS TL and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the aforementioned experiments, however.

The \texttt{PyTorch} machine learning framework is used, and many pre-trained models are taken from its \texttt{Torchvision} library. Exceptions are \texttt{EfficientNetV2}, \texttt{Swin}, \texttt{DeiT} and \texttt{BeiT}, which are taken from the \texttt{Timm} library\footnote{\url{https://timm.fast.ai/}.}.

Finally, all source code and data will be made available on GitHub\footnote{\url{https://github.com/IndoorAdventurer/ViTTransferLearningForArtClassification}.}.

>>>>

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pre-trained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that fine-tuning resulted in substantially better performance than either training from scratch or using an OTS learning scheme.

@@

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K, and the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used for the target tasks. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks extracted from it are: (1) \textit{Type classification}, where the model has to distinguish between classes as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with classes as `paper', `porcelain', `silver', etc.; and finally, (3) \textit{Artist classification}, where the model has to predict who the creator is. Figure \ref{methods:eg} shows labelled examples for all three tasks.

The full dataset contains 112,039 images, and allows for multiple or zero labels per classification task. Naturally, samples are excluded from a task if they contain no labels for it. Samples containing more than one label are excluded as well, since analysis showed that in all cases this is true for only a small portion of the dataset.% Note that this last point is different from \citeauthor{sabatelli2018deep}, which took the first occurring class in these cases.

@@

All datasets are split up into 80\%, 10\%, 10\% training, validating and testing sets. For robustness, five datasets are randomly generated per classification task, such that each experiment can be run five times. The average sample overlap between two of the five sets is shown in table \ref{methods:datasets} as well.

Besides the tasks described above, table \ref{methods:datasets} shows an experiment called \textit{Scaling}. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are gradually shrunk. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\% and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure that the distribution remains the same. Finally, also here 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

To start off with CNNs, \texttt{ResNet50} \citep{he2016deep} and \texttt{VGG19} \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for respectively FT and OTS learning. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used \texttt{ResNet50} too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used \texttt{ResNet101} and \texttt{ResNet152}. To also add some more recent CNNs, \texttt{ConvNext} \citep{liu2022convnet} -- a purely CNN-based model inspired by VTs' recent success -- and \texttt{EfficientNetV2} \citep{tan2021efficientnetv2} are included.

For VTs, versions with a 16x16 patch size are chosen. The first model is the original VT \citep{dosovitskiy2020image}, which will be referred to as \texttt{ViT}. Next, \texttt{Swin} \citep{liu2021swin} is considered, as it showed promising performance in \citeauthor{zhou2021convnets}. The final two are \texttt{DeiT} and \texttt{BeiT}. The small version of \texttt{DeiT} was used in \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}), because it is similar in size to \texttt{ResNet50}, to which it was compared. In this thesis, however, also a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.

@@

All VT-based models have a bit over 86 million parameters. \texttt{ConvNext} is of similar size, with 88.6 million parameters. \texttt{ResNet50} and \texttt{EfficientNetV2} are smaller, with 25.6 and 13.6 million, respectively, while \texttt{VGG19} is the largest model, with 143.7 million parameters.

@@

For OTS learning, the Adam optimizer \citep{kingma2014adam} is used with standard \texttt{PyTorch} parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for VT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32 GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS learning and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the aforementioned experiments, however.

The \texttt{PyTorch} machine learning framework is used \citep{paszke2019pytorch}, and many pre-trained models are taken from its \texttt{Torchvision} library. Exceptions are \texttt{EfficientNetV2}, \texttt{Swin}, \texttt{DeiT} and \texttt{BeiT}, which are taken from the \texttt{Timm} library\footnote{\url{https://timm.fast.ai/}.}.

Finally, all source code and data will be made available on \texttt{GitHub}\footnote{\url{https://github.com/IndoorAdventurer/ViTTransferLearningForArtClassification}.}.
